\section{Literature Review}
\label{sec:Lit}

Figure~\ref{fig:litmap} summarises how the sources in the research bibliography relate to conceptual themes. Branches group survey, runtime verification, automotive and trust, and methods/tooling; each citation is paired with a short summary (no duplicated sources).

\input{includes/literature-map}

This section reviews how recent studies approach verification, monitoring, and trust for connected, resource-constrained controllers such as automotive ECUs and IoT devices. The discussion is grounded in the sources charted in Fig.~\ref{fig:litmap}~\cite{williams2022survey,krichen2023survey,kuang2021remote,zhang2024context,dehoz2022iotdt,jaksic2018algebraic,saadat2024rosmonitoring2,clemens2018runtime,incki2018novel,mehran2024containers,goodloe2023highassurance,ekert2021cybersecurity,lorych2024hta,plappert2023hta,aman2020hatt,colombo2009larva,abela2023rvtee,saunders2016research}.

\subsection{Methodological themes in related work}

Several methodological strands recur across these studies. \textbf{Survey and taxonomy approaches} synthesise large bodies of technique: Krichen~\cite{krichen2023survey} maps formal verification and validation (FV\&V) methods for IoT, highlighting where exhaustive analysis gives way to lighter runtime checks, while Williams \textit{et al.}~\cite{williams2022survey} catalogue IoT security challenges under emerging connectivity paradigms. \textbf{Empirical and industry-oriented testing} is represented by Ekert \textit{et al.}~\cite{ekert2021cybersecurity}, who relate automotive cybersecurity V\&V to structured test campaigns and normative expectations. \textbf{Runtime monitoring with formal undertones} appears in Zhang \textit{et al.}~\cite{zhang2024context}, who use automata-theoretic environment models for context-aware safety monitoring in intelligent transport settings, and in de Hoz Diego \textit{et al.}~\cite{dehoz2022iotdt}, who combine an IoT digital twin with rule-based runtime verification over parametric network events, showing that observing only a subset of traffic can still expose protocol violations with low overhead. \textbf{Tool-chain and integration studies} include ROSMonitoring~2.0~\cite{saadat2024rosmonitoring2}, which extends runtime verification in ROS to services and ordered topics, and work on hardware trust anchors (HTA), where Lorych and Plappert~\cite{lorych2024hta} treat authentication for updatable IoT devices and Plappert \textit{et al.}~\cite{plappert2023hta} evaluate HTA designs against automotive requirements. Finally, Jak\v{s}i\'c \textit{et al.}~\cite{jaksic2018algebraic} advance an algebraic view of runtime verification, supporting incremental and quantitative monitoring---a useful reference when specifications are refined iteratively.

\subsection{Academic and non-academic material}

\textbf{Academic} sources here include peer-reviewed journals (e.g.~\cite{ekert2021cybersecurity,williams2022survey,plappert2023hta,zhang2024context,krichen2023survey}), archival conference proceedings (e.g.~\cite{lorych2024hta}), and edited volume chapters (e.g.~\cite{dehoz2022iotdt}). Workshop and electronic proceedings such as EPTCS~\cite{saadat2024rosmonitoring2} are scholarly but differ from flagship journals in review depth and indexing. \textbf{Non-academic or grey} material relevant to automotive work often comprises standards (e.g.\ ISO/SAE norms referenced in testing practice~\cite{ekert2021cybersecurity}), vendor security guides, and certification artefacts; these are essential for deployment context but are not substitutes for peer-reviewed evidence. The algebraic framework of Jak\v{s}i\'c \textit{et al.}~\cite{jaksic2018algebraic} is disseminated via arXiv: it is valuable for technical content but should be read as a preprint unless a later archival version is cited. Distinguishing these layers avoids overstating the evidential weight of grey literature while still acknowledging regulatory drivers.

\subsection{Five recommended peer-reviewed journal articles}

For readers prioritising archival journal literature, the following five articles offer strong overlap with automotive cybersecurity, IoT assurance, and runtime-oriented analysis: (1)~Ekert \textit{et al.}~\cite{ekert2021cybersecurity} on automotive cybersecurity V\&V testing; (2)~Krichen~\cite{krichen2023survey} on FV\&V for IoT; (3)~Williams \textit{et al.}~\cite{williams2022survey} on IoT security and emerging technologies; (4)~Plappert \textit{et al.}~\cite{plappert2023hta} on hardware trust anchors in automotive settings; (5)~Zhang \textit{et al.}~\cite{zhang2024context} on automata-theoretic, context-aware monitoring for autonomous vehicle safety. Together they span testing practice, formal-method surveys, security landscape, trust hardware, and concrete monitoring formalisms.

\subsection{Contextualisation for this study}

The present research targets a \emph{minimal, resource-bounded} runtime monitor on a safety-related ECU. In that framing, Ekert \textit{et al.}~\cite{ekert2021cybersecurity} contextualises the scale and structure of automotive security testing; Williams \textit{et al.}~\cite{williams2022survey} explains why connectivity increases exposure on constrained nodes; and Plappert \textit{et al.}~\cite{plappert2023hta} together with Lorych and Plappert~\cite{lorych2024hta} ground trust mechanisms that could protect monitors and keys where software-only solutions are insufficient. Zhang \textit{et al.}~\cite{zhang2024context} and de Hoz Diego \textit{et al.}~\cite{dehoz2022iotdt} motivate automata- and rule-based runtime checks with selective observation---aligned with limiting bus or signal instrumentation on an ECU. ROSMonitoring~2.0~\cite{saadat2024rosmonitoring2} illustrates how modern tool chains compose middleware events with external oracles, which parallels composing a simulated ECU feed with an external monitor even when the vehicle stack is not ROS-based. Jak\v{s}i\'c \textit{et al.}~\cite{jaksic2018algebraic} provides vocabulary for quantitative or multi-valued verdicts if requirements later go beyond Boolean pass/fail.

\subsection{Critical comparison and knowledge gaps}

The surveyed work \textbf{complements} yet \textbf{does not fully cover} an integrated ECU-local runtime verification path. Surveys~\cite{krichen2023survey,williams2022survey} compare techniques at scale but, by design, stop short of prescribing a minimal monitor set for a single braking or powertrain controller. Digital-twin and IoT network studies~\cite{dehoz2022iotdt} optimise traffic observation on networked devices, whereas an ECU simulation emphasises \emph{internal} state and sensor plausibility rather than packet subsets. Automata-based context monitoring~\cite{zhang2024context} assumes rich environment models; replicating that fidelity on a student-grade simulation may trade precision for tractability. ROS-oriented tooling~\cite{saadat2024rosmonitoring2} demonstrates composable monitoring, but automotive CAN signal semantics and homologation test cases are not identical to ROS topics. HTA work~\cite{plappert2023hta,lorych2024hta} addresses roots of trust and update authentication yet leaves open \emph{what} temporal properties an on-device monitor should enforce under strict CPU/RAM caps. The algebraic framework~\cite{jaksic2018algebraic} generalises monitoring semantics but does not resolve deployment questions for legacy ECU binaries.

\textbf{Knowledge gaps} salient for this project include: (i)~lack of published, reproducible pipelines that pair \emph{lightweight} ECU signal simulation with declarative monitors such as LARVA for automotive-grade properties; (ii)~fragmented guidance on calibrating thresholds (e.g.\ temperature, RPM) against both safety and cybersecurity abuse cases; (iii)~limited reporting on measured overhead when co-locating monitors with attestation or secure boot on class-A/B automotive microcontrollers. Addressing~(i)--(iii) experimentally would extend the literature beyond the surveyed sources while building directly on their methodological lessons.
