\section{Findings \& Discussion of Results}
\label{sec:Fin}

This section reports the experimental outcomes of the server-side ECU trace replay with LARVA instrumentation toggled on/off. Results are drawn from the benchmark session with tuned \texttt{loadMapMismatch} threshold (\texttt{|Load-MAPSource| > 60}), using 3 warmup runs and 10 measured runs per mode on the same machine and JVM conditions. The experiment assumes telemetry is transmitted from an IoT/ECU endpoint to a server-side verification API.

\subsection{Result presentation}

Two regimes were evaluated:
\begin{itemize}
    \item \textbf{Baseline regime:} clean trace (\texttt{20180713-home2mimos\_clean.csv}).
    \item \textbf{Abnormal regime:} injected trace (\texttt{20180713-home2mimos\_clean\_abnormal.csv}) with approximately 20\% violation rows across four flags.
\end{itemize}

Table~\ref{tab:rv-overhead} summarises core timing and throughput metrics. Standard deviation is reported with each average to capture run-to-run variability.
All elapsed-time values reported in this section correspond to full-trace processing per request (68{,}433 rows).

\begin{table}[ht]
    \centering
    \caption{Runtime verification overhead summary (10 measured runs per mode)}
    \label{tab:rv-overhead}
    \begin{tabular}{l|c|c}
    \textbf{Metric} & \textbf{Baseline} & \textbf{Abnormal} \\ \hline
    RV ON elapsed (ms) & 115.3 $\pm$ 6.04 & 819.3 $\pm$ 19.29 \\
    RV OFF elapsed (ms) & 111.5 $\pm$ 20.03 & 103.4 $\pm$ 1.71 \\
    RV ON throughput (rows/s) & 594{,}937.4 $\pm$ 30{,}089.0 & 83{,}567.3 $\pm$ 1{,}940.6 \\
    RV OFF throughput (rows/s) & 626{,}961.4 $\pm$ 82{,}253.9 & 661{,}990.5 $\pm$ 10{,}913.9 \\
    Overhead (\%) & 3.41 & 692.36 \\
    \end{tabular}
\end{table}

Tail latency also reflects this split behaviour. For baseline, RV ON p95/p99 are 124.40/127.28 ms, while RV OFF p95/p99 are 145.75/161.95 ms. For abnormal traces, RV ON p95/p99 are 850.80/856.56 ms, whereas RV OFF remains at 106.00/106.00 ms.

\subsection{Analysis and interpretation}

\textbf{RQ1 (detection of injected abnormalities).} The abnormal regime produced high RV ON flag totals (\texttt{coolantTooHigh}=34210, \texttt{voltageOutOfRange}=34200, \texttt{throttleSpike}=34220, \texttt{loadMapMismatch}=34230), while RV OFF flags remained zero. This indicates that violation reporting is coupled to monitor execution as intended and that injected abnormal classes are detected consistently.

\textbf{RQ2 (monitoring overhead).} Baseline overhead is low (3.41\%), suggesting that monitor integration is feasible for non-adversarial trace processing on the tested server path. Abnormal overhead remains substantial (692.36\%), showing that violation-heavy execution paths still dominate runtime even after reducing logging overhead. In other words, monitor cost is not constant; it is strongly state-dependent. These overhead values should be interpreted as server-side verification cost under a telemetry-upload architecture.

\textbf{RQ3 (stability and threshold sensitivity).} Standard deviations are small relative to means in most cases, indicating repeatable behaviour. After tuning \texttt{loadMapMismatch} from 30 to 60, baseline false positives reduced (aggregate baseline \texttt{loadMapMismatch} decreased compared with pre-tuning runs), while abnormal detection remained strong. This supports threshold calibration as a practical control for false-positive pressure.

\subsection{Controlled root-cause experiment}

To isolate the source of abnormal-path overhead, an additional controlled comparison was executed on three monitor variants under identical benchmark settings (1 warmup, 5 measured runs): (i) original generated monitor, (ii) logging disabled, and (iii) logging disabled plus removal of bad-state stacktrace/string construction. Timing results are summarised in Table~\ref{tab:rv-overhead-variants}, and resource metrics are summarised in Table~\ref{tab:rv-resource-variants}.

\begin{table}[ht]
    \centering
    \caption{Abnormal-path overhead across monitor variants (controlled 3-way comparison)}
    \label{tab:rv-overhead-variants}
    \scriptsize
    \begin{tabular}{l|c|c|c}
    \textbf{Variant} & \textbf{Base O/H (\%)} & \textbf{Abn O/H (\%)} & \textbf{Abn RV ON (ms)} \\ \hline
    Original & 6.43 & 3927.05 & 4824.4 \\
    Logging disabled & 14.61 & 683.79 & 851.2 \\
    No stacktrace & 6.75 & 9.39 & 125.8 \\
    \end{tabular}
\end{table}

This progression shows that the dominant abnormal overhead originates from generated diagnostic behaviour in bad-state handling (especially stacktrace-backed state-string construction), not from the core guard checks themselves.

\begin{table}[ht]
    \centering
    \caption{Resource metrics across monitor variants (controlled 3-way comparison, abnormal runs)}
    \label{tab:rv-resource-variants}
    \scriptsize
    \resizebox{\columnwidth}{!}{%
    \begin{tabular}{l|c|c|c|c}
    \textbf{Variant} & \textbf{On CPU} & \textbf{Off CPU} & \textbf{On Heap $\Delta$} & \textbf{Off Heap $\Delta$} \\ \hline
    Original & 3562.5 & 125.0 & -30.59 & 1.85 \\
    Logging disabled & 850.0 & 100.0 & -10.62 & 12.19 \\
    No stacktrace & 125.0 & 112.5 & 13.97 & -26.13 \\
    \end{tabular}%
    }
\end{table}

CPU time follows the same trend as elapsed time, confirming that compute spent inside monitor-side diagnostics is the dominant contributor in abnormal paths. Heap delta is less stable across variants because Java GC activity during full-trace requests can produce negative or positive net deltas; therefore, heap delta is interpreted as a coarse indicator, while CPU and elapsed-time comparisons provide the primary evidence.

\subsection{Comparison with related work and quality considerations}

The observed trade-off aligns with prior RV literature: runtime monitoring provides actionable safety/cybersecurity visibility, but overhead depends on property design, trigger frequency, and deployment context~\cite{krichen2023survey,zhang2024context,jaksic2018algebraic}. The clear difference between baseline and abnormal runtime also mirrors broader automotive V\&V concerns, where worst-case and abuse-case behaviour must be measured rather than assumed from nominal conditions~\cite{ekert2021cybersecurity}.

\textbf{Validity.} Internal validity is strengthened by same-endpoint comparisons, warmup before measured runs, and on/off toggling as the only intended treatment factor. Construct validity is supported by explicit metrics (elapsed time, throughput, p95/p99, overhead) and deterministic replay inputs.

\textbf{Reliability.} Repeated 10-run measurements with reported standard deviations indicate reproducibility under stable local conditions.

\textbf{Generalizability.} External validity remains limited to this server-side replay implementation and selected property set. Results should not be extrapolated directly to production ECUs without hardware-in-the-loop or in-vehicle evaluation.